### 2015-07-02 ApTranscriptome_TR.Rmd

```{r setup, echo=FALSE, results='hide', message = FALSE, warning=FALSE}
# Global settings
options(stringsAsFactors=FALSE, digits = 3, scipen = 1)

@@

In this technical report, which accompanies the manuscript **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species** (Stanton-Geddes et al., in press), we:

1. Describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`.
2. Identify thermally-responsive genes.
3. Evaluate differences in transcriptome-wide patterns of thermal responsiveness (i.e. the thermal reactionome) between the two species
4. Perform gene set enrichment analysis of thermally-responsive genes for the two species.

@@

Per suggestion of a reviewer, make a cartoon figure that shows the contrasting results of the *enhance defense* and the *tolerance* hypotheses.

The *enhanced response hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response. This pattern would be apparent as an excess of transcripts with the same response pattern between species (e.g. more shared "Low" transcripts) and a deficit of transcripts that shift to other response patterns or become not responsive.

```{r ed_cartoon}
enhanced_defense_dat <- md_sub1
enhanced_defense_dat$value <- 0
# set non-shared response to random negative value 
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low"), "value"] <- runif(5, min=140, max=150)
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal"), "value"] <- rep(-200, 5)
# change shared response to positive
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low" & enhanced_defense_dat$Apic == "Low"), "value"] <- 50
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal" & enhanced_defense_dat$Apic == "Bimodal"), "value"] <- 50

enhanced_defense_dat

# plot
enhanced_defense_plot <- qplot(y=Apic, x=Acar, data=enhanced_defense_dat, fill=value, geom="tile", xlim = rev(levels(enhanced_defense_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "right") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

# title
enhanced_defense_plot <- enhanced_defense_plot + ggtitle('Enhanced Response hypothesis') + theme(plot.title = element_text(hjust=0, size = 12))
enhanced_defense_plot
```

In contrast, the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species.

@@

# plot
tolerance_plot <- qplot(y=Apic, x=Acar, data=tolerance_dat, fill=value, geom="tile", xlim = rev(levels(tolerance_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

# title
tolerance_plot <- tolerance_plot + ggtitle('Tolerance hypothesis') + theme(plot.title = element_text(hjust=0, size = 12))
tolerance_plot

# plot together
png("results/predictions_Fig1.png")
grid.arrange(enhanced_defense_plot, tolerance_plot, ncol=2, widths = c(3.75,3))
dev.off()
```

The number of thermally-responsive in each response category differs between the colonies, with the msot transcripts expressed at *Low* temperatures in both colonies. For *ApVT*, an equal number of transcripts are expressed at *High* and *Bimodal*, followed by *Intermediate* transcripts. For *AcNC*, transcripts expressed at *Intermediate* temperatures are next most common, followed by *High* and *Bimodal*.

@@

An interesting observation from the matched observations plot (Fig. 1) is that for **NotResp** genes in *A. carolinensis* the **High** and **Bimodal** categories in *A. picea* are over-represented. Finding that the mean expression level of these genes in *A. carolinensis* is greater than the expression level at the optimum temperature (19.5°C) in *A. picea* would be consistent with genetic assimilation.

@@

Next, I check the same for the **Bimodal** genes in *A. picea*.

@@

As with the **High** transcripts, expression much higher for the responsive *A. picea* transcripts than for the non-responsive *A. carolinensis* transcripts.

@@

```{r responsive_pred}
# apply predFunc to all responsive transcripts
resp.TPM.dt.sub.pred <- ddply(resp.TPM.dt.sub, .(Transcript), .inform="TRUE", predFunc)

@@

To assist visualization of specific transcripts, I made a interactive web-app using the [shiny](http://www.rstudio.com/shiny/) package. The scripts for this app are in the sub-directory `.\ApRxN-shinyapp`.

@@

save(sigtable, chi1, chi2, chi3, chi4, chi5, chi6, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, hsp_responsive, hsp_all, mh.test, mh_plot, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, md, file = paste("doc/ms_slim_", Sys.Date(), ".RData", sep = ""))
```

>>>>

```{r setup, echo=FALSE, results='hide', message = FALSE, warning=FALSE}
# Global settings
options(stringsAsFactors=FALSE, digits = 3, scipen = 1)
options(width = 80)

@@

In this technical report, which accompanies the manuscript **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species**, we:

1. Describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`.
2. Measure the reaction norm across a thermal gradient (0°C to 38.5°C) for all transcripts and characterize the *thermal reactionome* as the set of transcripts that are thermally-responsive.
3. Quantitatively test three mechanistic hypotheses of thermal adapation:.
  - The *Enhanced response* hypothesis proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors.
  - The *Tolerance hypothesis* proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency.
  - The *Genetic assimilation hypothesis* proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes.

We quantify the relative importance of each of these mechanisms to thermal adaptation in these two species by evalutating differences in the reactionomes between species. We use gene set enrichment analysis to provide additional insight on the cellular processes underlying the transcriptome-wide changes in expression.

@@

The *enhanced response hypothesis* posits that temperature adaptation uses existing response mechanisms, which should result in significant overlap in response. This pattern would be apparent as an excess of transcripts with the same response pattern between species (e.g. more shared "Low" transcripts) and a deficit of transcripts that shift to other response patterns or become not responsive.

```{r er_cartoon}
enhanced_response_dat <- md_sub1
enhanced_response_dat$value <- 0
# set non-shared response to random negative value 
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low"), "value"] <- runif(5, min=140, max=150)
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal"), "value"] <- rep(-200, 5)
# change shared response to positive
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low" & enhanced_defense_dat$Apic == "Low"), "value"] <- 50
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal" & enhanced_defense_dat$Apic == "Bimodal"), "value"] <- 50

enhanced_response_dat

# plot
enhanced_response_plot <- qplot(y=Apic, x=Acar, data=enhanced_response_dat, fill=value, geom="tile", xlim = rev(levels(enhanced_response_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "right") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

enhanced_response_plot <- ggplot(enhanced_response_dat, aes(y = Apic, x = Acar, fill=value)) + 
  geom_tile(colour = "#ffffff", size =1) +
  xlim(rev(levels(enhanced_response_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "right") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

# title
enhanced_response_plot <- enhanced_response_plot + ggtitle('Enhanced Response hypothesis') + theme(plot.title = element_text(hjust=0, size = 10))
enhanced_response_plot
```

In contrast, the *tolerance hypothesis* predicts that genes involved in active response will become non-responsive or shift to other response categories in the better-adapted species.

@@

# plot
tolerance_plot <- ggplot(tolerance_dat, aes(y=Apic, x=Acar, fill=value)) + 
  geom_tile(colour = "#ffffff", size =1) +
  xlim(rev(levels(enhanced_response_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

# title
tolerance_plot <- tolerance_plot + ggtitle('Tolerance hypothesis') + theme(plot.title = element_text(hjust=0, size = 10))
tolerance_plot

# plot together
png("results/predictions_Fig1.png")
grid.arrange(enhanced_response_plot, tolerance_plot, ncol=2, widths = c(3.7, 3))
dev.off()
```

The number of thermally-responsive in each response category differed between the colonies, with the msot transcripts expressed at *Low* temperatures in both colonies. For *ApVT*, an equal number of transcripts are expressed at *High* and *Bimodal*, followed by *Intermediate* transcripts. For *AcNC*, transcripts expressed at *Intermediate* temperatures are next most common, followed by *High* and *Bimodal*.

@@

An interesting observation from the matched observations plot (Fig. 1) is that for **NotResp** genes in *A. carolinensis* the **High** and **Bimodal** categories in *A. picea* are over-represented. Finding that the mean expression level of these genes in *A. carolinensis* is greater than the expression level at the rearing temperature (25°C) in *A. picea* would be consistent with genetic assimilation.

@@

Next, I checked the same for the **Bimodal** genes in *A. picea*.

@@

As with the **High** transcripts, expression was much higher for the responsive *A. picea* transcripts than for the non-responsive *A. carolinensis* transcripts.

@@

```{r responsive_pred}
# subset to transcripts with significant temperature x species interaction
resp.TPM.dt.sub <- TPM.dt.sub[names(interaction.lms)]

# apply predFunc to all responsive transcripts
resp.TPM.dt.sub.pred <- ddply(resp.TPM.dt.sub, .(Transcript), .inform="TRUE", predFunc)

@@

To assist visualization of specific transcripts, I made a interactive web-app using the [shiny](http://www.rstudio.com/shiny/) package. The scripts for this app are in the sub-directory `.\ApRxN-shinyapp`. The web app is available at https://johnsg.shinyapps.io/ApRxN-shinyapp/

@@

save(sigtable, chi1, chi2, chi3, chi4, chi5, chi6, resp.TPM.dt.sub, Ap.type.table, Ap.type.table.prop, tt2, tt3, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, hsp_responsive, hsp_all, mh.test, mh_plot, T_on.high.ttest, T_on.low.ttest, T_off.int.high.ttest, T_off.int.low.ttest, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, Ar.int.A22.lowbim.gsea, md, responsive_transcripts_w_BLAST, annotation.table.responsive, file = paste("doc/ms_slim.RData", sep = ""))
```
